Algorithmic reliability is an important aspect in many scientific and engineering fields.
Despite the advancements in computational genetics, the focus has mostly been on performance while sacrificing the provided guarantees.
We resolve the long-standing trade-off between optimality and performance for the read to graph mapping problem.
Here we formally define the problem of read to graph mapping and present a provably-optimal practical solution.
We developed Probabilistic Variant Graph (PVG) – a generative stochastic model designed for read mapping to general reference graphs.
It completely encompasses all of the various available data sources in their probabilistic interpretation:
	1) read quality scores, 2) edit distance alignment metric, and 3) variant likelihood statistics.
This model captures an unprecedented level of generality while guaranteeing global seed-and-extend mapping optimality.
Our novel heuristic precomputation allows an A* shortest path algorithm to reach a near input-size empirical complexity.
We demonstrate that aligning using our model improves the mapping accuracy of current state of the art tools while matching the performance on both synthetic and real data.
Because of being a generative model, we are able to optimize the alignment probability P(read|path) but also to estimate the marginal likelihood and mapping probabilisty P(path|read).
The PVG tool is available on https://github.com/eth-sri/pvg.

%\textbf{Contact:} \href{pesho@inf.ethz.ch}{pesho@inf.ethz.ch}\\
%\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics} online.
