\section{Overview}

We present an algorithm for the \emph{optimal alignment} of sequences to
\emph{genome graphs}. It works by phrasing the edit distance minimization task
as finding a shortest path on an implicit alignment graph. To find a shortest
path, we instantiate the \A paradigm with a novel domain-specific heuristic
function that accounts for the upcoming subsequence in the query to be aligned,
resulting in a provably optimal alignment algorithm called \astarix.
Experimental evaluation of \astarix shows that it is 1--2 orders of magnitude
faster than state-of-the-art optimal algorithms on the task of aligning Illumina
reads to reference genome graphs.

All optimal algorithms have to iterate over the whole reference for each query.
We should be able to optimize this by precomputation.

\paragraph{Methods}
We introduce a novel approach, called \astarix, for optimal sequence-to-graph
alignment based on \A. As with any \A instantiation, the core difficulty lies in
developing an accurate domain-specific heuristic which is fast to compute. We
design a heuristic that accounts for the content of the upcoming query letters
to be aligned, which more effectively guides the search. Our proposed heuristic
has two advantages: (i)~it is correctness-preserving, that is, it preserves the
fact that \astarix finds the best alignment, yet (ii)~it is practically
effective in that the algorithm performs a near-optimal number of steps.
Overall, this heuristic enables \astarix to compute the best alignment while
also scaling to larger reference graph sizes when compared to existing
state-of-the-art optimal aligners.

\paragraph{Results}
% From outside
%In the context of semi-global sequence-to-graph alignment, \A has been used to
%empirically scale sublinearly with the reference size for short reads.

\begin{samepage}
The main contributions presented in this chapter are:
	
\begin{enumerate}
	\item \textbf{\astarix.} An algorithm for optimal sequence-to-graph
	alignment based on a novel instantiation of \A with an accurate
	domain-specific heuristic that accounts for the upcoming query letters to be
	aligned (\cref{TRIEsec:astarix-algo}).
	\item \textbf{Algorithmic optimizations.}
	To ensure that \astarix is practical, we introduce a number of algorithmic
	optimizations which increase performance and decrease memory footprint
	(\cref{TRIEsec:optimizations}). We also prove that all optimizations are
	correctness-preserving.
	\item \textbf{Thorough experimental evaluation of \astarix.}
	We demonstrate that \astarix is up to 2 orders of magnitude faster than
	other optimal aligners on various reference graphs (\cref{TRIEsec:eval}).
\end{enumerate}
\end{samepage}
