%\section{Summary}

We presented \astarix, an \A algorithm to find optimal alignments, based on a
domain-specific heuristic and enhanced by multiple algorithmic optimizations.
Importantly, our approach allows for both cyclic and acyclic graphs including
variation and de Bruijn graphs. We demonstrated that \astarix scales exponentially better than \dijkstra with
increasing (but small) number of errors in the reads. Moreover, for short reads,
both \astarix and \dijkstra scale better and outperform current state-of-the-art
optimal aligners with increasing genome graph size. Nevertheless, scaling
optimal alignment of long reads on big graphs remains an open problem.

We have presented an optimal read aligner based on the \A~algorithm instantiated
with a novel \seedh which guides the search by preferring crumbs on nodes that lead
towards optimal alignments even for long reads.

We presented an algorithm with an implementation in \astarpa solving pairwise alignment between
two sequences. The algorithm is based on \A with a \sh, inexact matching, match
chaining, and match pruning, which we proved to find an exact solution according
to edit distance. For random sequences with up to $15\%$ uniform errors, the
runtime of \astarpa scales near-linearly to very long sequences ($10^7\bp$)
and outperforms other exact aligners. We demonstrate that on real ONT reads from
a human genome, \astarpa is faster than other aligners on only a limited portion of the
reads.

% from the grand proposal
We initiated a new direction of sequence alignment based on the informed
shortest path algorithms A* that we have shown to be both provably optimal,
practically scalable and more performant than existing aligners in certain
cases. The objective of this project is to develop our work on A* for sequence
alignment from prototypical to practical, which includes the development of
algorithms, formal proofs and software development. Our specific aim is to use
the already constructed base for the development of AStarix as an
industrial-scale mapper handling human-scale graph references, long and noisy
reads, while being competitive in runtime and memory even to suboptimal
aligners. In order to fulfill this aim, work in several directions will be
needed: 1) rigorous development of the seed heuristic theory, 2) theoretical
analysis of the expected runtime, 3) implementation optimized for runtime,
memory, parallelization and use-cases. Our vision is that A* could become the
default approach for sequence alignment, useful to the computational biology,
bioinformatics and the more general algorithmic community.

%2.5. Relevance and impact
The potential adoption of the \A approach to sequencing (incl. the usage of
AStarix mapper and the A*PA global aligner) may hugely alleviate the
decades-long conflict between alignment optimality and computational costs. This
basic task is an extremely common element in the upstream pipelines and other
high-level tools like assemblers. Possibly, many alignment tools can be
outcompeted (similar to the Parasail aligner) and become outdated. The general
impact of guaranteed optimality is increased trust in the tools by
bioinformaticians, ability to explain any alignment, reduced number of
misalignments, applicability to both linear and graph references. A risky
high-stake problem we plan to explore is the multiple sequence alignment (MSA)
since it has many resemblances to pairwise alignment but suffers from the
combinatorial explosion with the number of sequences. Even with relaxed
optimality-guarantees, a novel approach to this general problem may have a
direct impact on comparative genomics, including protein folding. Our experience
with presenting the \A approach convinced us that it is not a common knowledge
in the computation biology community, even though most conference attendants
with technical background are familiar with the name and the general usage. Our
conviction is that \A is a powerful instrument that has fallen out of focus in
our community but that may prove useful for other algorithmic biology problems
as well.Potentially, it could become a standard topic in the university
bioinformatics courses.

% paper:global
Our graph-based approach to alignment differs considerably from dynamic
programming approaches, mainly because of the ability to use information from the
entire sequences. This additional information enables radically more focused
path-finding at the cost of more complex algorithms.
