% intuitive
\section*{Background}
\addcontentsline{toc}{section}{\protect\numberline{}{Background}}

%\paragraph{Context}
%\paragraph{DNA discovery}
The discovery of the DNA structure and function\citep{watson1953structure}
seventy years ago was a revolution in our understanding of the living world.
Subsequently, the technologies for reading (\emph{sequencing}) of genetic data
exploited the non-branching (linear or circular) DNA structure, similar to human
text and speech. Since then, sequence alignment has been a core problem in
computational biology with a multitude of applications. In this thesis I propose
a novel principled approach to this problem, and demonstrate its superior
scaling and performance on a variaty of genomic data.

\paragraph{First generation sequencing}
Genomic (DNA, RNA) sequencing is a technology that takes genetic material and
produces a data file with a set of sequences. Each sequence from a new
biological sample can be compared to a reference genome (if such is available)
in order to identify mutations, quantify expression per gene, and many other
types of analysis. To figure out the location in the reference each sequence
most probably originates from, it should first be semi-globally aligned to the
reference. If the sequences of two proteins or viruses have been reconstructed,
aligning them globally reveals the most probabale mutations between them.

%\paragraph{Applications}

\paragraph{Second generation (high-throughput) sequencing}
Initially, the sequenced data has been scarce and the sequence alignment was
done by hand. The development of high-throughput sequencing in the late 90s and
2000s lead to higher amounts of cheaper data due to sequencing millions and
billions of molecules in parallel. The amount and complexity of the available
genomic data bacame not only intractable by hand, but also challenging for
computer analysis.With the advance of the sequencing technology, the need for
fast and accurate algorithms has been steadily
increasing~\cite{alser2021technology}. One possible reason is the combination of
massive new data and advancement in computation hardware allowing for various
kinds of parallelization.
%Thus the advancements with bit-parallel algorithms (CITE Myears, Mikko).

\paragraph{Third generation (long-read) sequencing}
Since the end of the 2000s, novel technologies started to appear that are
capable of producing long sequences. Long sequences enabled massive applications
but the limiting factor was the high error rate (up to $20\%$) and the higher
price.

\paragraph{Metagenomics and genome graphs}
% graph references for semi-global
The abundance of data enabled the assembly of richer references that capture
biological variation from different individuals and species. In particular, the
last decade saw the increased development of graph references that compactly
represent a whole set of genomes (a metagenome) as paths. Semi-global alignment
of de novo sequences on such graphs produces more accurate alignments than on
linear references.

% paper-trie; Accounting for variation
%\paragraph{Accounting for Variation}
First attempts to include variation into the reference data structure were made
by augmenting the local alignment method to consider alternative walks during the
extend step~\cite{schneeberger_simultaneous_2009,palmapper}. This approach has
since been extended from the linear reference case to graph references. To
represent non-reference variation of multiple references during the seeding
stage, HISAT2 uses generalized compressed suffix
arrays~\cite{siren_indexing_2014} to index walks in an augmented reference
sequence, forming a local genome graph~\cite{kim_graphbased_2019}.
VG~\cite{garrison_variation_2018} uses a similar
technique~\cite{siren_indexing_2017} to index variation graphs representing a
population of references.

% paper-trie; The benefit from genome graphs
Historically, a single linear reference sequence has been used to represent the
most common variants in a population. While providing a working abstraction for
most cases, rare or sub-population specific variation is especially hard to
model in this setting, creating a reference allele
bias~\cite{stevenson_sources_2013,brandt_mapping_2015}. Consequently, in the
last few years, the field has shifted first towards using sets of reference
sequences, and more recently to graph data structures (so-called {\em genome
graphs}), to represent many genomes or haplotypes
simultaneously~\cite{dilthey_improved_2015,paten_genome_2017,garrison_variation_2018}.
