% This text will be sanitized and placed into lqa-output/abstract.txt
%
Motivation and approach: the trie index works for scaling sublinerly with the
reference size but each error in the query triggers deeper exploration of the
trie so the runtime grows exponentially. Lets inform the \A algorithm using
information from the whole query length while keeping the trie. This way we
could avoid the deep trie exploration and scale to long queries (as far as the
error rate is not too high).

We present a novel \A \textit{\seedh} that enables fast and optimal
sequence-to-graph alignment, guaranteed to minimize the edit distance of the
alignment assuming non-negative edit costs.

We phrase optimal alignment as a shortest path problem and solve it by
instantiating the \A~algorithm with our \seedh. The \seedh first extracts
non-overlapping substrings (\textit{seeds}) from the read, finds exact seed
\textit{matches} in the reference, marks preceding reference positions by
\textit{crumbs}, and uses the crumbs to direct the \A search. The key idea is to
punish paths for the absence of foreseeable seed matches. We prove admissibility
of the \seedh, thus guaranteeing alignment optimality.

\qquad Our implementation extends the free and open source aligner and
demonstrates that the \seedh outperforms all state-of-the-art optimal aligners
including \graphaligner, \vargas, \pasgal, and the \prefixh previously employed
by \astarix. Specifically, we achieve a consistent speedup of >60$\times$ on
both short Illumina reads and long HiFi reads (up to 25kbp), on both the
\textit{E.~coli} linear reference genome (1Mbp) and the MHC variant graph
(5Mbp). Our speedup is enabled by the \seedh consistently skipping >99.99\% of
the table cells that optimal aligners based on dynamic programming
compute.\\

\astarix aligner and evaluations: \astarixurl\\

Genome graph, Optimal alignment, Semi-global alignment, Edit distance, Shortest
path, Long reads, \A algorithm, Seed heuristic
