\subsection{Match pruning preserves finding a shortest path}\label{GLOBALsec:pruning-proof}

\begin{lem}\label{GLOBALprop:consistent}
  For both $\hshS$ and $\hcshS$\,, any expanded state at the start of a seed is fixed.
\end{lem}
\begin{proof}
  Let $\hh$ be either $\hshS$ or $\hcshS$.
  We use a proof by contradiction.
  Thus, suppose that a state $u$ at the start of a seed has minimal $f$ among
  all open states, but the shortest path $\pi^*$ from $v_s$ to $u$ has length $\g(u) < g(u)$.

  Let $n\neq u$ be the last expanded state on $\pi^*$, and let $v$ be its successor.
  Let the unexpanded states of $\pi^*$ following $n$ that are at the start of a seed be
  $v\preceq w_0 \prec \dots\prec w_l=u$, in this order. We will show that
  $f(v) < f(u)$, resulting in a contradiction with the fact that \A always
  expands the open state with minimal $f$:
  \begin{align*}
    f(v) &= g(v)+\hh(v)
         = \g(v)+\hh(v) && \text{$\pi^*$ is a shortest path to $v$,} \\
         &\leq \g(w_0)+\hh(w_0) && \text{proven in part 1,} \\
         &\phantom{aaa}\leq \cdots \leq \g(w_l)+\hh(w_l) && \text{proven in part 2,} \\
         &= \g(u)+\hh(u)
         < g(u)+\hh(u) = f(u) && \text{by contradiction.}
  \end{align*}

  \textbf{Part 1}: Proof of $\g(v)+h(v) \leq \g(w_0)+h(w_0)$.
  Since $w_0$ is the first start of a seed at or after $v$, the set of seeds
  following $v$ is the same as the set of seeds following $u$. Thus, the
  summation in $\hh(v)$ and $\hh(w_0)$ in \cref{GLOBALeq:sh} or \cref{GLOBALeq:csh} is over
  the same seeds. Since $v\preceq w_0$, this implies $\hh(v) \leq \hh(w_0)$ in
  both cases, and $\g(v) \leq \g(w_0)$ since $v\preceq w_0$.

  \textbf{Part 2}: Proof of $\g(w_i){+}\hh(w_i) \leq \g(w_{i+1}){+}\hh(w_{i+1})$.
  Note that $w_i$ and $w_{i+1}$ are at the start of the same or consecutive
  seeds. In the first case,
  $\hh(w_i) \leq \hh(w_{i+1})$ follows by definition since $w_i \preceq w_{i+1}$.
  Else, we have $\hh(w_i) \leq \hh(w_{i+1})+r$.
  When $d:=d(w_i, w_{i+1}) < \spot$, there is an unpruned match $m$ from $w_i$
  to $w_{i+1}$ in $M-E$ of cost $d$, since $w_i$ is not expanded.
  This implies $\hh(w_i) \leq \hh(w_{i+1}) + d$, and we obtain
  \begin{align*}
    \g(w_i) + \hh(w_i) &= (\g(w_{i+1}) - d) + \hh(w_i)\\
                       &\leq (\g(w_{i+1})-\min(d,r)) + (\hh(w_{i+1}) + \min(d,r))\\
                       &= \g(w_{i+1}) + \hh(w_{i+1}).
  \end{align*}
  Thus, by contradiction, $\g(u) < g(u)$ for any expanded start of a seed $u$.
\end{proof}

\begin{lem}\label{GLOBALlem:partialadmissible}
  The heuristics $\hcshS$ and $\hshS$ are partially-admissible.
\end{lem}
\begin{proof}
  Let $\pi^*$ be a shortest path from $s$ to $v_t$. Let $n$ be the last
  expanded state on $\Path$ at the start of a seed.
  By \cref{GLOBALprop:consistent} $n$ is fixed.
  By choice of $n$, no states on $\pi^*$ following $n$ are expanded, so
  no matches on $\pi^*$ following $n$ are pruned.
  The proof of \cref{GLOBALthm:admissible} applies to the path $\Path$ without
  changes, implying that $h(u) \leq \h(u)$ for all $u$ on $\Path$ following
  $n$, for both $h=\hshS$ and $h=\hcshS$.
\end{proof}

\thmprunedcorrect
\begin{proof}
Follows from~\cref{GLOBALthm:admissible} in~\cref{GLOBALsec:partial-admissible}
and~\cref{GLOBALlem:partialadmissible}.
\end{proof}

